Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Guideline for International \emph{and Domestic} Students

### 2024-02-27 demystify.tex
updates

Ultimately, choose a prof. that fits you the most by communicating with them, taking their courses, meeting and asking them questions, and even talking to their current students. It will take time and effort, but remember, you will be working with this person for 5+ years, so it is important to try to find the right one.

@@

\paragraph{Tenure-line and teaching faculty} Tenure line faculty, consisting of tenured and tenured-track profs., focus on research, which includes publishing papers, obtaining grants, mentoring Ph.D students.  They often have very low teaching load (e.g., 1 per semester). \emph{Tenured} faculty are professors who have been promoted to have a permanent position (informally, very hard to fire them).   \emph{Tenure-track} faculty are (more likely young) professors who are on the track to get tenure.  \textbf{Both tenure and tenured track faculty can serve as a formal adviser of Ph.D. students}. This is very important because you need to have a tenure line faculty as your main PhD adviser. \autoref{sec:tenure-vs-tenure-track} talks more about choosing between tenure and tenured-track professor as your adviser.

In contrast, \emph{teaching} (or instructional faculty or professor of practice) mainly focus on teaching. They typically teach 3--4 classes per semester (which is quite a lot) and \emph{do not} have research responsibilities (e.g., they do not have to worry much about publishing papers or obtaining grants).  They can mentor Ph.D. students but they typically cannot serve as a formal adviser of Ph.D. students (i.e., they \emph{cannot graduate} PhD students).

It is worth noting that professors in a non-CS department also unlikely can serve as formal PhD advisers for CS students (see \autoref{sec:related-fields}).

\paragraph{Titles} Faculty have rankings, e.g., assistant, associate, full, regardless if they are tenure line or teaching.  Assistant means new faculty, associate means they have been promoted, and full means they are senior. A tenure track faculty starts with being an assistant professor and then gets tenure and promoted to associate (typically after 6 years) and then full (time varies, some become full within 3-4 years, some 10+ years, some remain associate).

@@

Tenured faculty, e.g., associate and full profs., are likely older, more well-known, and have more experience in managing students.  However, they might not push you as hard (they don't have to, they already got tenure). They might also expect you to figure things out yourself, i.e., so you need to be very independent.  Some tenured faculty are also no longer active in research and more involved with administrative responsibilities or with their startup companies (this means they will likely not take new students).

@@

Faculty received many "cold" e-mails from international students seeking for admission, TA, and RA. Most of the time, we ignore these emails, but in some rare occasions we do answer them. So why do we ignore some but reply to others and how to write an email that get our attention?

@@

If you want to contact a prof. to ask about \emph{research opportunities}, or \emph{GTA/GRA} support, then \emph{yes}, I believe you should---it is \emph{worth it}. However, you need to put effort into it and really do it the \emph{right way}.

@@

\emph{MS Degree requirement}:  as mentioned in \autoref{sec:msrequirement} and \ref{sec:time}, CS PhD programs in the US do not require MS degree.  In contrast, many other countries do require MS degree before joining a PhD program.  This means that US PhD programs are longer (5--7 years, 2 of which are coursework) than other countries (3--4 years, no coursework).

>>>>

Ultimately, choose a prof. that fits you the most by communicating with them, taking their courses, meeting and asking them questions, and talking to their current students. It will take time and effort, but since you will be working with this person for 5+ years, it is important to try to find the right one.

@@

\paragraph{Tenure-line and teaching faculty} Tenure line faculty, consisting of tenured and tenured-track profs., focus on research, which includes publishing papers, obtaining grants, and mentoring Ph.D students.  They often have very low teaching load (e.g., 1 per semester). \emph{Tenured} faculty are professors who have been promoted to have a permanent position (informally, very hard to fire them).   \emph{Tenure-track} faculty are (more likely young) professors who are on the track to get tenure.  \textbf{Both tenure and tenured track faculty can serve as a formal adviser of Ph.D. students}. This is very important because you need to have a tenure line faculty as your main PhD adviser. \autoref{sec:tenure-vs-tenure-track} talks more about choosing between tenure and tenured-track professor as your adviser.

In contrast, \emph{teaching} (or instructional faculty or professor of practice) mainly focus on teaching. They typically teach 3--4 classes per semester (which is quite a lot) and \emph{do not} have research responsibilities, e.g., they do not have to worry much about publishing papers or obtaining grants. They can mentor Ph.D. students but they typically cannot serve as a formal adviser of Ph.D. students, i.e., they \emph{cannot graduate} PhD students.

It is worth noting that professors in a non-CS department unlikely can serve as formal PhD advisers for CS students (see \autoref{sec:related-fields}).

\paragraph{Titles} Faculty have titles such as assistant, associate, full, regardless if they are tenure line or teaching.  Assistant means new faculty, associate means they have been promoted, and full means they are senior. A tenure track faculty starts with being an assistant professor and then gets tenure and promoted to associate (typically after 6 years) and then full (time varies, some become full within 3-4 years, some 10+ years, some remain associate). A non-tenure track faculty, e.g., teaching faculty or lecturer, also starts with assistant, then promoted to associated, and full.

@@

Tenured faculty, e.g., associate and full profs., are likely older, more well-known, and have more experience in managing students.  However, they might not push you as hard. They might also expect you to figure things out yourself, i.e., so you need to be independent.  Some tenured faculty are also no longer active in research and more involved with administrative responsibilities or with their startup companies (this means they will likely not take new students).

@@

Faculty received many "cold" e-mails from prospective students seeking for admission, TA, and RA. Most of the time, we ignore these emails, but in some rare occasions we do answer them. So why do we ignore some but reply to others and how to write an email that get our attention?

@@

If you want to contact a prof. to ask about \emph{research opportunities}, or \emph{GTA/GRA} support, then \emph{yes}, I believe you should---it is \emph{worth it}. However, you really need to put effort into it and do it the \emph{right way}.

@@

\emph{MS Degree requirement}:  as mentioned in \autoref{sec:msrequirement} and \autoref{sec:time}, CS PhD programs in the US do not require MS degree.  In contrast, many other countries do require MS degree before joining a PhD program.  This means that US PhD programs are longer (5--7 years, 2 of which are coursework) than other countries (3--4 years, no coursework).
